---
title: "P8106_HW3_sp4436"
author: "Sukhman Parhar"
date: "2026-04-01"
output: html_document
---

```{r, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ISLR)
library(glmnet)
library(caret)
library(tidymodels)
library(corrplot)
library(ggplot2)
library(plotmo)
library(ggrepel)
library(pls)
library(pdp)
library(earth)
library(splines)
library(mgcv)
library(patchwork)
library(readr)
library(pROC)
library(vip)
library(discrim)
library(MASS)
library(klaR)
```

#### Clean the data

```{r, warning = FALSE, message = FALSE}
# Load in the data 
auto <- read_csv("auto.csv") %>% 
  na.omit() %>% 
  as.data.frame()

# Remove cars with 3 cylinders
auto <- auto[auto$cylinders != 3, ]

# Convert Outcome (mpg_cat) to factor
auto$mpg_cat <- factor(auto$mpg_cat)

# Convert Origin to factor 
auto$origin <- as.factor(auto$origin)

#summary(auto)
```

#### Split the data

```{r}
set.seed(2)

split <- initial_split(auto, prop = 0.7)

train_data <- training(split)
test_data  <- testing(split)

# set up cross validation control
ctrl <- trainControl(method = "cv", number = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)
```

#### Question 1: Perform logistic regression analysis. Are there redundant predictors in your model? If so, identify them. If there are none, please provide an explanation.

```{r}
# Fit tuning GLM 
glm.fit <- glm(mpg_cat ~ ., data = train_data, family = binomial)
summary(glm.fit)
```

```{r}
set.seed(2)

model.glm <- train(mpg_cat ~ ., 
                   data = train_data,
                   method = "glm",
                   metric = "ROC",
                   trControl = ctrl)
# Summary
summary(model.glm)

print(model.glm)
```

###

Yes there are redundant predictors within this model. Only weight and year are statistically significant, while cylinders, displacement, horsepower, and acceleration have high p-values. These predictors are redundant because they measure  aspects of engine size and performance. Year and origin capture distinct information and are not redundant.

<hr style="height:2px;border-width:0;color:black;background-color:black">

#### Question 2: Train a multivariate adaptive regression spline (MARS) model. Does the MARS model improve prediction performance compared to logistic regression?

```{r, warning = FALSE, message = FALSE}
# MARS
set.seed(2)

mars_grid <- expand.grid(degree = 1:3, 
                         nprune = 2:20)

model.mars <- train(mpg_cat ~ ., 
                   data = train_data,
                    method = "earth",
                    tuneGrid = mars_grid,
                    metric = "ROC",
                    trControl = ctrl)

# Best tuning parameters for MARS
model.mars$bestTune

# Plot CV performance for MARS
plot(model.mars)

# compare cross-validation performance
res <- resamples(list(Logistic = model.glm,
                      MARS = model.mars))

summary(res)

bwplot(res, metric = "ROC", main = "CV ROC: Logistic vs. MARS")
```

###
The MARS model showed better performance, with a higher mean ROC (0.974955 vs. 0.970505), higher sensitivity (0.949451 vs. 0.928022), and higher specificity (0.947253 vs. 0.910440). MARS improves predictive performance compared to logistic regression.

<hr style="height:2px;border-width:0;color:black;background-color:black">

#### Question 3: Perform linear discriminant analysis using the training data. Plot the linear discriminant(s).

```{r}
# Fit LDA model
lda.fit <- lda(mpg_cat ~ ., data = train_data)

# View model
lda.fit

plot(lda.fit)

lda.fit$scaling
```

```{r}
set.seed(2)

model.lda <- train(mpg_cat ~ ., 
                   data = train_data,
                   method = "lda",
                   metric = "ROC",
                   trControl = ctrl)
```

<hr style="height:2px;border-width:0;color:black;background-color:black">

#### Question 4: Which model will you choose to predict the response variable? Plot its ROC curve and report the AUC.

```{r}
set.seed(2)

# Compare all models
res <- resamples(list(
  Logistic = model.glm,
  MARS = model.mars,
  LDA = model.lda
))

summary(res)
```

#### MARS performed the best 

```{r}
# Predict probabilities (MARS)
mars.probs <- predict(model.mars,
                      newdata = test_data,
                      type = "prob")[,2]

# ROC curve
roc.mars <- roc(test_data$mpg_cat, mars.probs)

# AUC
auc(roc.mars)

# Plot ROC
plot(roc.mars, legacy.axes = TRUE,
     main = "ROC Curve - MARS Model")
```

###
Area under the curve: 0.9795

<hr style="height:2px;border-width:0;color:black;background-color:black">

#### Question 5: Select a probability threshold to classify observations and compute the confusion matrix for the final model you choose. Briefly interpret what the confusion matrix indicates about your model’s performance

```{r}
# Predict probabilities High
test.probs <- predict(model.mars, newdata = test_data, type = "prob")

test.pred.prob <- test.probs[, "high"]

# Convert 
test.pred <- rep("low", length(test.pred.prob))
test.pred[test.pred.prob > 0.5] <- "high"

# confusion matrix
confusionMatrix(data = as.factor(test.pred), 
                reference = test_data$mpg_cat, 
                positive = "high")
```

###
A  threshold of 0.5 was selected because choosing 0.5 assumes that the costs associated with False Positives (misclassifying a low-efficiency car as high) and False Negatives (failing to identify a high-efficiency car) are equal.

Mars has an overall Accuracy of 0.9316. 
The Sensitivity of 0.9483 means it can identify true high-mileage vehicles, while the Specificity of 0.9153 shows it can identify low-mileage vehicles. 